---
title: "Climbing Sim"
output: rmdformats::robobook 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      fig.width = 6,
                      fig.height = 7)
```


```{r}
library(tidyverse)
theme_set(theme_light())
```

# Qualification

Note: 20 athletes participate in qualification. 8 advance to final round.

Let's try 10000 sims for now.

```{r}
set.seed(1)
nsim <- 10000
qual_sim <- list()
for (i in 1:nsim) {
  qual_sim[[i]] <-
    bind_cols(
      player = 1:20,
      e1 = sample(1:20, replace = FALSE),
      e2 = sample(1:20, replace = FALSE),
      e3 = sample(1:20, replace = FALSE)
    ) %>%
    mutate(sim = i)
}

qual <- bind_rows(qual_sim) # tidyverse's do.call-rbind version

qual <- qual %>% 
  mutate(qual_score = e1 * e2 * e3) %>% 
  group_by(sim) %>% 
  arrange(qual_score, .by_group = TRUE) %>% 
  mutate(qual_rank = row_number()) 
```

## If you win the first event, what is the probability that you advance?

```{r}
qual %>% 
  filter(e1 == 1 & qual_rank <= 8) %>% 
  ungroup() %>% 
  summarize(nrow(.)/nsim)
```

## If you win ANY event, what is the probability you advance?  

```{r}
qual %>% 
  filter((e1 == 1 | e2 == 1 | e3 == 1) & qual_rank <= 8) %>% 
  nrow()
```

## If you finish last in the first event, what is the probability of advancing? 

```{r}
qual %>% 
  filter(e1 == 20 & qual_rank <= 8) %>% 
  nrow()
```

## If you finish last in ANY event, what is the probability you advance? 

```{r}
qual %>% 
  filter((e1 == 20 | e2 == 20 | e3 == 20) & qual_rank <= 8) %>% 
  nrow()
```

## What is the score that guarantees that you advance / do not advance? 


## What is the average score of someone who advances? 

```{r}
qual %>% 
  filter(qual_rank <= 8) %>% 
  ungroup() %>% 
  group_by(qual_rank) %>% 
  summarize(avg_adv_score = mean(qual_score))
```

```{r}
qual %>% 
  filter(qual_rank <= 8) %>% 
  ungroup() %>% 
  summarize(avg_adv_score_overall = mean(qual_score))
```


## What is/are the distribution on the the first/second/etc. place qualifier? 

```{r}
qual %>% 
  ggplot(aes(x = factor(qual_rank), y = qual_score)) +
  geom_dotplot(binaxis = "y", binwidth = 1) + 
  coord_flip()
```

```{r}
library(ggridges)
qual %>% 
  ggplot(aes(x = qual_score, y = factor(qual_rank))) +
  geom_density_ridges()
```

# Final Round

```{r}
set.seed(1)
nsim <- 10000
final_sim <- list()
for (i in 1:nsim) {
  final_sim[[i]] <-
    bind_cols(
      player = 1:8,
      e1 = sample(1:8, replace = FALSE),
      e2 = sample(1:8, replace = FALSE),
      e3 = sample(1:8, replace = FALSE)
    ) %>%
    mutate(sim = i)
}

final <- bind_rows(final_sim) 

final <- final %>% 
  mutate(final_score = e1 * e2 * e3) %>% 
  group_by(sim) %>% 
  arrange(final_score, .by_group = TRUE) %>% 
  mutate(final_rank = row_number()) 
```

## If you win any event, what is the probability that you win a medal?

```{r}
final %>% 
  filter((e1 == 1 | e2 == 1 | e3 == 1) & final_rank <= 3) %>% 
  nrow()
```

## If you win the first event, what is the probability you win a medal?

```{r}
final %>% 
  filter(e1 == 1 & final_rank <= 3) %>% 
  nrow()
```

## What is the distribution of the winners score if ranks are random? 

```{r}
final %>% 
  filter(final_rank == 1) %>% 
  ggplot(aes(final_score)) + 
  geom_histogram(bins = 21, color = "white")
```
